---
title: "GROUP_THREE_ICL"
author: "Group Three"
date: "`r Sys.Date()`"
output: html_document
---

```         
```

## PLAN

-   Import the dataset (clinic_vital - excel file)

-   Cleaned summary to spot data entry problems

    -   data entry problems to understand patient population before health outreach campaign

        -   cleaning columns : replacing missing by means

        -   calculations : BMI

    -   Data Visualizations

    -   Interepretation

## Importing the dataset

```{r}
# import excel
clinic_vitals <- read_excel("~/Documents/data_group_work/data/03_clinic_vitals.xlsx")

# display first 10 rows
head(clinic_vitals)
```

### Inspecting the data

```{r}
# show data information : length, class, mode
summary(clinic_vitals)
```

```{r}
# display data structure : columns , datatype , length
str(clinic_vitals)
```

```{r}
# check length and columns : length 1024 rows, columns (9)
dim(clinic_vitals)
```

#### Check for impossible age values

```{r}
# sum of ages below 0 and more than 100 (close to 200)
# less than zero or more than 100 
sum(clinic_vitals$age < 0 | clinic_vitals$age > 100)
```

```{r}
# replacing ages below zero and above 100 with NA (missing)
clinic_vitals$age[clinic_vitals$age < 0 |clinic_vitals$age > 100 ] <- NA
```

```{r}
# checking changes applied
table(clinic_vitals$age)
```

```{r}
# checking changes applied : length and columns
dim(clinic_vitals)
```

#### check for missing values in weight_kg and height_m

```{r}
# sum of missing values 
colSums(is.na(clinic_vitals))
```

```{r}
# replacing missing values of weight with mean
clinic_vitals$weight_kg[is.na(clinic_vitals$weight_kg)] <- mean(clinic_vitals$weight_kg, na.rm =  TRUE)
```

```{r}
# replacing missing values of height with mean
clinic_vitals$height_m[is.na(clinic_vitals$height_m)] <- mean(clinic_vitals$height_m, na.rm =  TRUE)
```

```{r}
# sum of missing values 
colSums(is.na(clinic_vitals))
```

### Systolic_bp and Diastolic_bp check

```{r}
# checking count of where sys_bp < dia_bp
sum(clinic_vitals$systolic_bp < clinic_vitals$diastolic_bp, na.rm = TRUE)
```

```{r}
# swapping occurrence where the systolic < diastolic
# step 1: assigned and index to cases where sys_bp < dias_bp
index <- clinic_vitals$systolic_bp < clinic_vitals$diastolic_bp

# store dias_index in a store 
temporary_store <- clinic_vitals$diastolic_bp[index]

# swap 
clinic_vitals$systolic_bp[index] <- clinic_vitals$diastolic_bp[index]

# retrieve
clinic_vitals$diastolic_bp[index] <- temporary_store

```

```{r}
# checking count of where sys_bp < dia_bp
sum(clinic_vitals$systolic_bp < clinic_vitals$diastolic_bp, na.rm = TRUE)
```

### Cleaning the smoker column (Yes/NO)

```{r}
# cleaning from yes,y to Yes 
clinic_vitals$smoker[tolower(clinic_vitals$smoker) %in% c("yes", "y")] <- "Yes"
clinic_vitals$smoker[tolower(clinic_vitals$smoker) %in% c("no", "n")] <- "No"
```

```{r}
head(clinic_vitals$smoker, 100)
```

### Duplicated Rows

```{r}
# identifying the count of duplicated rows 
sum(duplicated(clinic_vitals))
```

```{r}
# identifying the duplicated rows themselves
clinic_vitals[duplicated(clinic_vitals), ]
```

```{r}
clinic_vitals <- clinic_vitals[!duplicated(clinic_vitals), ]
```

```{r}
# identifying the count of duplicated rows 
sum(duplicated(clinic_vitals))
```

## Calculating BMI

### BMI = weight / (height \*\* 2)

```{r}
clinic_vitals$BMI <- clinic_vitals$weight_kg / clinic_vitals$height_m ** 2
```

## Visualisations

### Histogram

```{r}
# frequency distribution 
hist(clinic_vitals$BMI, main = "Body Mass Index (BMI) DISTRIBUTION", xlab = "Body Mass Index (BMI)", ylab = "Count of people", col = "blue")
```

## Histogram Interpretation

-   The histogram shows the distribution of Body Mass Index (BMI) among the individuals in the dataset.

-   The x-axis represents BMI values, while the y-axis represents the frequency (number of individuals) within each BMI range.

-   The bars indicate how frequently different BMI values occur.

-   The BMI distribution appears approximately normally distributed, with most individuals having BMI values concentrated around the center and fewer individuals at the lower and higher extremes.

-   This suggests that the BMI values in the study population are fairly evenly distributed around the average.

### Boxplot

```{r}
# plotting the boxplot of systolic bp against clinical_vitals
boxplot(clinic_vitals$systolic_bp ~ clinic_vitals$smoker, main = "Systolic BP by Smokers", xlab = "Smoker", ylab = "Systolic BP", col = "blue")
```

### Boxplot Interpretation

-   The non-smoker group has several low outliers, with systolic BP values below 80 mmHg. These observations are unusually low compared with the rest of the non-smoker group.

-   The upper whisker extends to approximately 155 mmHg for non-smokers and 150 mmHg for smokers, suggesting that both groups have individuals with relatively high systolic BP.

-   The middle 50% of observations (IQR) for both groups is approximately between 115 and 130mmHg hence a less variability

### Scatter plot

```{r}
# plotting the scatter plot of systolic bp against smokers 
plot(clinic_vitals$age, clinic_vitals$systolic_bp, main = "Age vs Systolic Bp", xlab = "Age (years)" , ylab =" Systolic BP", col = "blue")
```

### Scatterplot interpretation

-   The points are widely scattered, indicating that systolic BP varies considerably among individuals of the same age.

-   Most systolic BP values are concentrated approximately between 110 and 130, although some values are lower or higher.

-   There is a weak linear relationship between age and systolic BP in this data set since there is no pattern.

-   Age alone does not appear to be a strong predictor of systolic blood pressure based on the pattern of this scatter plot.

```{r}
plot.new()

text(0.5, 0.6, "THANK YOU",
     cex = 3,
     font = 2)

text(0.5, 0.45, "Thank you for your attention!",
     cex = 1.5)
```
